Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-28 astronum_2018/astronum-xrb.tex
Suggested reworking of section 3.
Try to link each potential improvement to its impact on number of
zones, or timesteps, or computational cost per zone.

Add some back-of-the-envelope estimates of the benefits where
possible.

The 2D calculations shown in the previous section require about 10k
node-hours of computational time to get to about 10~ms (those
particular simulations were run at NERSC).  Most of this time is spent
in the reactions\MarginPar{verify fraction}.  We really need to do 3D.
There the shear from the hurricane structure acting along the flame
front may cause instabilities that can affect the flame propagation.
This is something we can test with our current simulation methodology.
A 3D version of this calculation is estimated to require 5M node-hours
on CPUs---this is about the size of an annual \MarginPar{double check
  \#s} allocation.  Additionally, this is still with the boosted
flame, but we expect that once the flame is developed, the boosting
will not be as critical, due to hydrodynamic effects.  We plan to test
this soon.  We can bootstrap the 3D calculation by mapping the 2D
axisymmetric calculation, once the flame is firmly established, into
3D, rotating about the symmetry axis.  This will cut down the
computational time by about one-half.  In any case, this is less than
1\% of the neutron star surface, so a fully resolved calculation of
burning over the entire neutron star is out of reach, and will be for
some time.

Algorithmic improvements are needed to push the scale of simulations
to the point where we can model XRBs at the full star scale with
resolved nuclear physics.  Here we discuss a few possibilities.

@@

Another way we can meet the resolution requirements is to use a more
accurate hydrodynamic method.  We are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions. \MarginPar{need some refs here} Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames, and since computational work scales like
$(\Delta x)^4$ in 3D, this can save a lot of computer time.

@@

If we continue to resolve the flame structure, then we need to be
smarter (and more aggressive) about our grid refinement strategy.
Currently, we put the entire fuel layer on the finest grid, but for
the material ahead of the flame, we could keep it at lower resolution
until the flame reaches it.  The main obstacle is that we need to keep
the fuel ahead of the flame in hydrostatic equilibrium, and keeping it
refined accomplishes this well.  Using well-balanced
schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium
(HSE) in the lower refined regions, reducing the number of zones
needed to maintain fidelity, but we would still need a way to regrid
the layer ahead of the flame to higher resolution to accurately
capture the burning dynamics.  To accomplish this, we envision
building a refinement strategy that does interpolation of the data onto
the newly refined grids enforcing hydrostatic equilibrium.  This can
allow us to keep only a narrow fully refined region that ``slides''
along the surface of the neutron star keeping pace with the advancing
flame.  Such a scheme would greatly reduce the computational costs.
This is a longer term development.

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of XRBs, one possibility for
multiscale modelling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a low Mach number or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the 2D shallow water model
would be negligible compared to those of the low Mach/compressible model, the overall cost of
this multiscale model would be determined by the size of the low Mach/compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
An example of coupling compressible and low Mach number models implemented using the \amrex~framework, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implemenation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.

>>>>

The 2D calculations shown in the previous section require about 10k
node-hours of computational time to get to about 10~ms (those
particular simulations were run at NERSC).  We expect qualitatively different
results in 3D calculations: for example, the shear from the hurricane structure
acting along the flame front may cause instabilities that can affect the flame
propagation.
With our current simulation methodology,
a 3D version of this calculation is estimated to require 5M node-hours
on CPUs---this is about the size of an annual \MarginPar{double check
  \#s} allocation.
%Additionally, this is still with the boosted flame, but we expect that once the flame is developed, the boosting will not be as critical, due to hydrodynamic effects.
We plan to test this soon.
This simulation covers less than
1\% of the neutron star surface, so a fully resolved calculation of
burning over the entire neutron star is out of reach: a naive scaling
calculation suggest that 15 years of Moore's Law improvements are required.
%
% The calculation above is
% log_2 [ (4 pi 10^2) / (1.23^2) ] * 1.5
% for a NS of 10km radius vs the computational domain of 1.23km^2
%

Improvements are needed to push the scale of simulations
to the point where we can model XRBs at the full star scale with
resolved nuclear physics.
As the computational cost is roughly a product of the number of zones, the
number of zone updates, and the cost per zone update, we need to look for
improvements in each of these. The number of zones depends on the physical
model employed, the meshing strategy, and the accuracy required. The number of
zone updates depends on the smallest zone size and the largest characteristic
speed in the model. The computational cost per zone update is dominated by the
time spent in the reactions\MarginPar{verify fraction}. We discuss a few
possible improvements that can come both from implementing new algorithms and
from exploiting new hardware.

%For example, the number of zones can be reduced
%by symmetry considerations. We can bootstrap the 3D calculation by mapping the
%2D axisymmetric calculation, once the flame is firmly established, into
%3D, rotating about the symmetry axis.  This will cut down the computational
%time by about one-half.

@@

Another way we can reduce the number of zones for given accuracy is to use a more
accurate hydrodynamic method.  We are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions. \MarginPar{need some refs here} Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames. For smooth wave solutions we may expect to need
around half the number of zones per dimension, and since computational work
scales like $(\Delta x)^4$ in 3D, this could lower costs by a factor of around
10.
%
% This calculation is based on the number of points per wavelength required,
% to resolve a periodic wave, which (I believe) goes from 12 (2nd order) to
% about 7 (4th order). However, this is very idealized, and I'm not sure that
% it carries over to the reactions, so this may well be optimistic.
%

@@

If we continue to resolve the flame structure, then we need to be
smarter (and more aggressive) about our grid refinement strategy.
Currently, we put the entire fuel layer on the finest grid, but for
the material ahead of the flame, we could keep it at lower resolution
until the flame reaches it.  The main obstacle is that we need to keep
the fuel ahead of the flame in hydrostatic equilibrium, and keeping it
refined accomplishes this well.  Using well-balanced
schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium
(HSE) in the lower refined regions, reducing the number of zones
needed to maintain fidelity, but we would still need a way to regrid
the layer ahead of the flame to higher resolution to accurately
capture the burning dynamics.  To accomplish this, we envision
building a refinement strategy that does interpolation of the data onto
the newly refined grids enforcing hydrostatic equilibrium.  This can
allow us to keep only a narrow fully refined region that ``slides''
along the surface of the neutron star keeping pace with the advancing
flame.  When using mesh refinement the overall costs are dominated by the
number of zones at the finest resolution, so such a scheme would greatly reduce
the computational costs. This is a longer term development.

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of XRBs, one possibility for
multiscale modelling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a low Mach number or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the 2D shallow water model
would be negligible compared to those of the low Mach/compressible model, the overall cost of
this multiscale model would be determined by the size of the low Mach/compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
As with the more aggressive gridding strategy above, the hybrid approach
reduces the number of zones with the largest computational cost through mesh
refinement, and also reduces the computational costs where possible by
employing cheaper physical models in regions where this is possible.
An example of coupling compressible and low Mach number models implemented using the \amrex~framework, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implemenation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.
